# Comparing several means (one-way ANOVA).

**Learning objectives:**

- What ANOVA is
- Build ANOVA from scratch
- testing available functions 


## Introduction

This is an introduction to the widely used statistical tool, the `Analysis of Variance (ANOVA)`. Initially developed by Sir Ronald Fisher in the early 20th century, the term ANOVA is somewhat misleading as it primarily deals with investigating differences in means, not variances. 

The chapter focuses on the simplest form of ANOVA, known as `one-way ANOVA`, applicable when dealing with multiple groups of observations to discern variations in an outcome variable.



## Example: Clinical Trial
```{r}
library(lsr)
load(here::here("data/clinicaltrial.RData"))
str(clin.trial)
```
```{r}
clin.trial
```

We look at the `effect of drug on mood.gain`

```{r}
library(tidyverse)
clin.trial%>%
  group_by(drug)%>%
  reframe(avg_mood.gain=mean(mood.gain),
          sd_mood.gain=sd(mood.gain))
```
```{r}
gplots::plotmeans(formula = mood.gain ~ drug,  # plot mood.gain by drug
             data = clin.trial,           # the data frame
             xlab = "Drug Administered",  # x-axis label
             ylab = "Mood Gain",          # y-axis label
             n.label = FALSE  )
```

## Use ANOVA


The Null hypothesis is:
$$H_0: \text{it is true that } \mu_P=\mu_A=\mu_J$$
While the alternative is:

$$H_0: \text{it is NOT true that } \mu_P=\mu_A=\mu_J$$


The Sample Variance of Y:
$$Var(Y)=\frac{1}{N}\sum_{k=1}^G\sum_{i=1}^{N_k}(Y_{ik}-\bar{Y})^2$$
### Example 2

```{r}
N<- 5 # number of people
G<- 2 # groups
```


```{r}
data2 <- tibble(name=c("Ann","Ben","Cat","Dan","Egg"),
       person_p=seq_along(1:5),
       group=c("cool","cool","cool","uncool","uncool"),
       group_k=c(1,1,1,2,2),
       index_i=c(1,2,3,1,2),
       grumpiness_Yp=c(20,55,21,91,22)
       )
data2
```
$$Var(Y)=\frac{1}{N}\sum_{p=1}^{N}(Y_{p}-\bar{Y})^2$$



## The sum of squares

$$SS_{tot}=\sum_{k=1}^G\sum_{i=1}^{N_k}(Y_{ik}-\bar{Y})^2$$
### Between-group sum of squares
$$SS_{b}=\sum_{k=1}^G\sum_{i=1}^{N_k}(Y_{k}-\bar{Y})^2$$
$$=\sum_{k=1}^G{N_k}(\bar{Y}_{k}-\bar{Y})^2$$
$$SS_w+SS_b=SS_{tot}$$
## The F-test

$$df_b=G-1$$
$$df_w=N-G$$
$$MS_b=\frac{SS_b}{df_b}$$
$$MS_w=\frac{SS_w}{df_w}$$

$$F=\frac{MS_b}{MS_w}$$


## Example3

```{r}
data3 <- tibble(group_k=c("placebo","placebo","placebo","anxifree","anxifree"),
       outcome_Yk=c(0.5,0.3,0.1,0.6,0.4),
       group_mean=c(0.45,0.45,0.45,0.72,0.72),
       mean_dev=outcome_Yk-group_mean,
       sqr_dev=mean_dev^2)
data3
```
```{r}
ss_w <- sum(data3$sqr_dev)
ss_w
```
```{r}
outcome <- clin.trial$mood.gain
group <- clin.trial$drug
tibble(outcome,group)
```


```{r}
# ?tapply
gp.means <- tapply(outcome,group,mean)
```


```{r}
gp.means <- gp.means[group]
```


```{r}
dev.from.gp.means <- outcome - gp.means
squared.devs <- dev.from.gp.means ^2
```
```{r}
Y <- clin.trial%>%
  group_by(drug)%>%
  reframe(mood.gain,
          gp.means=mean(mood.gain),
          dev.from.gp.means=mood.gain-gp.means,
          squared.devs=(mood.gain-gp.means)^2,
          sample_size=rep(6,length(drug)))

Y
```
```{r}
ssw<-sum(Y$squared.devs)
ssw
```

```{r}
grand_mean <- mean(Y$mood.gain)
```


```{r}
data4 <- Y%>%
  group_by(drug)%>%
  reframe(group_mean=mean(mood.gain))%>%
  mutate(grand_mean=grand_mean,
         deviation=group_mean-grand_mean,
         sqr_dev=deviation^2,
         sample_size=c(6,6,6),
         w_sqr=sample_size*sqr_dev)
          
data4
```


```{r}
ssb <- sum(data4$w_sqr)
ssb
```

```{r}
ssw;ssb
```
```{r}
G <- 3
N<- 18
dfb <- G-1
dfw <- N-G
```


```{r}
msb <- ssb/dfb
msw <- ssw/dfw
```


F(2,15) = 18.6 
```{r}
f <- msb/msw
f
```

?pf()

```{r}
pf( f, df1 = 2, df2 = 15, lower.tail = FALSE)
```

Reject the NULL hypothesis:
```{r}
pf( f, df1 = 2, df2 = 15, lower.tail = FALSE) < 0.05
```
## Testing available functions

?aov()

```{r}
my.anova <- aov( formula = mood.gain ~ drug, 
                 data = clin.trial ) 
my.anova
```
```{r}
names(my.anova)
```

```{r}
summary(my.anova)
```


```{r}
posthocPairwiseT( my.anova )
```

## Apply a correction for multiple comparisons

###  Bonferroni corrections

The corrected p-value is the result of multiply all your raw p-values by m, where m is the number of separate tests. Such as in the previous case we had to compare placebo vs drug1, placebo vs drug2, drug1 vs drug2; so we had 3 tests. 

$${p}'=p*m$$
${p}'< \alpha = 0.05$


```{r}
posthocPairwiseT( my.anova, 
                  p.adjust.method = "bonferroni")
```


### Holm corrections

Another method, pretending the tests are done sequentially. 

$${p}'_j=j*p_j$$

```{r}
posthocPairwiseT( my.anova )
```

### Normality, Homogeneity of variance and Independence

#### Welch one-way test
F(2,)
```{r}
oneway.test(mood.gain ~ drug, data = clin.trial)
```

```{r}
oneway.test(mood.gain ~ drug, data = clin.trial, var.equal = TRUE)
```

## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
<summary> Meeting chat log </summary>

```
LOG
```
</details>
